---
title: "MC1"
author: "Min Htet Aung"
description: "This is the Take-home Exercise 2" 
date: "May 15, 2025"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Loading Packages

```{r}
pacman::p_load(tidyverse, jsonlite,SmartEDA, tidygraph,ggraph,igraph)
```

# Reading jason file

```{r}
kg <- fromJSON("data/MC1_graph.json")
```

# Inspect structure

```{r}
str(kg,max.level = 1)
```

## Convert JSON to DataFrames

```{r}
nodes_tbl <- as_tibble(kg$nodes)
edges_tbl <- as_tibble(kg$links)
```

## Explore and Clean the Data

```{r}
glimpse(nodes_tbl)
glimpse(edges_tbl)
```

## Map Node IDs to Index for Graph Conversion

```{r}
id_map <- tibble(id = nodes_tbl$id, index = seq_len(nrow(nodes_tbl)))

edges_tbl <- edges_tbl %>%
  left_join(id_map, by = c("source" = "id")) %>%
  rename(from = index) %>%
  left_join(id_map, by = c("target" = "id")) %>%
  rename(to = index) %>%
  filter(!is.na(from), !is.na(to))
```

## Initial EDA

### Count of Edge Types

```{r}
ggplot(data = edges_tbl,
       aes(y = `Edge Type`)) +
  geom_bar()
```

### Count of Node Types

```{r}
ggplot(data = nodes_tbl,
       aes(y = `Node Type`)) +
  geom_bar()
```



## Creating Knowledge Graph

### Step 1: Mapping form node id to row index


id_map <- tibble(id = nodes_tbl$id, 
                 index = seq_len(
                   nrow(nodes_tbl)
                 ))


### Step 2 : Map source and target IDs to row indices


edges_tbl <- edges_tbl %>%
  left_join(id_map, by = c("source" = "id")) %>%
  rename(from = index) %>%
  left_join(id_map, by = c("target" = "id")) %>%
  rename(to = index)


### Step 3: filter out any unmatched (invalid) edges


edges_tbl <- edges_tbl %>%
  filter(!is.na(from), !is.na(to))


### step 4: Creating the graph


graph <- tbl_graph(nodes = nodes_tbl,
                   edges = edges_tbl,
                   directed = kg$directed)


## Visualising the Knowledge graph


set.seed(1234)


### Visualising the whole Graph


#| eval: false
ggraph(graph, layout = "fr") + 
  geom_edge_link(alpha = 0.3,
                 colour = "gray") +
  geom_node_point(aes(color = `Node Type`),
                  size = 4) +
  geom_node_text(aes(label = name),
                 repel = TRUE,
                 size = 2.5) +
  theme_void()


#### Step 1: Filter edges to only "Member of"


graph_memberof <- graph %>%
  activate(edges) %>%
  filter(`Edge Type` == "MemberOf")


#### Step 2: Extract only connected nodes (i.e., used in these edges)


used_node_indices <- graph_memberof %>%
  activate(edges) %>%
  as_tibble() %>%
  select(from, to)%>%
  unlist() %>%
  unique()


#### Step 3: Keep only those nodes


graph_memberof <- graph_memberof %>%
  activate(nodes) %>%
  mutate(row_id = row_number()) %>%
  filter(row_id %in% used_node_indices) %>%
  select(-row_id)


#### Plot the sub-graph


ggraph(graph_memberof,
       layout = "fr") + 
  geom_edge_link(alpha = 0.5,
                 colour = "gray") +
  geom_node_point(aes(colour = `Node Type`),
                  size = 1) +
  geom_node_text(aes(label = name),
                 repel = TRUE,
                 size = 2.5) +
  theme_void()

