---
title: "Hand-on Exercise 4.2"
author: "Min Htet Aung"
description: "This is the Hands-on Exercise 4.2" 
date: "May 7, 2025"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installing and Launching R packages

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```


## Importing data

```{r}
exam <- read_csv("data/Exam_data.csv")
```


## One-sample test: `gghistostats()` method

```{r}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```


## Two-sample mean test: `ggbetweenstats()`

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE
)
```


## Oneway ANOVA Test: `ggbetweenstats()` method

```{r}
ggbetweenstats(
  data = exam,
  x = RACE, 
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```


## Significant Test of Correlation: `ggscatterstats()`

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE,
  )
```


## Significant Test of Association (Depedence) : `ggbarstats()` methods

```{r}
exam1 <- exam %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)
```

In this code chunk below `ggbarstats()` is used to build a visual for Significant Test of Association

```{r}
ggbarstats(exam1, 
           x = MATHS_bins, 
           y = GENDER)
```
