---
title: "Hands-on Exercise 8.2"
author: "Min Htet Aung"
description: "This is the Hands-on Exercise 8.2" 
date: "Jun 09, 2025"
date-modified: "last-modified"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Visualising Geospatial Point Data

## Importing Packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```


# Geospatial Data Wrangling

## Data Import and Preparation

The code chunk below uses `read_csv()` function of readr package to import `SGPools_svy21.csv` into R as a tibble data frame called sgpools.

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

```{r}
list(sgpools) 
```


## Creating a sf data frame from an aspatial data frame

The code chunk below converts sgpools data frame into a simple feature data frame by using `st_as_sf()` of sf packages

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

```{r}
list(sgpools_sf)
```


# Drawing Proportional Symbol Map

To create an interactive proportional symbol map in R, the view mode of tmap will be used.

The code churn below will turn on the interactive mode of tmap.

```{r}
tmap_mode("view")
```


## It all started with an interactive point symbol map

The code chunks below are used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```


## Lets make it proportional

To create a proportional symbol map, the `Gp1Gp2Winnings` variable is mapped to the size aesthetic.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```


## Lets give it a different colour

The proportional symbol map can be further improved by using the colour visual attribute. In the code chunks below, OUTLET_TYPE variable is used as the colour attribute variable.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```


## Side by side visualization

tmap’s view mode supports faceted plots using `tm_facets(sync = TRUE)` to enable synchronized zoom and pan across maps.

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) + 
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

```{r}
tmap_mode("plot")
```