---
title: "Hands-on Exercise 2"
author: "Min Htet Aung"
description: "This is the Hands-on Exercise 2" 
date: "April 24, 2024"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started 


```{r}
pacman::p_load(ggrepel, patchwork, ggthemes, hrbrthemes, tidyverse)
```


## Importing data

```{r}
exam_data <- read.csv("Exam_data.csv")
```


# Beyond ggplot2 Annotation: ggrepel

## Working without ggrepel

```{r}
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method = lm, size=0.5) + geom_label(aes(label = ID), hjust = .5, vjust = -.5) + coord_cartesian(xlim = c(0,100), ylim = c(0,100)) + ggtitle("English Scores Vs Maths Scores for Primary 3")
```


## Working with ggrepel

```{r}
ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method = lm, size=0.5) + geom_label_repel(aes(label = ID), fontface = "bold") + coord_cartesian(xlim = c(0,100), ylim = c(0,100)) + ggtitle("English Scores Vs Maths Scores for Primary 3")
```


# Beyond ggplot2 Themes

ggplot2 have built-in themes - 
- **theme_gray()**
- **theme_bw()** 
- **theme_classic()**
- **theme_dark()**
- **theme_light()**
- **theme_linedraw()**
- **theme_minimal()**
- **theme_void()**

## Using ggplot2 themes

```{r}
ggplot(data=exam_data, aes(x=MATHS)) + geom_histogram(bins = 20, boundary = 100, color = "grey25", fill = "grey90") + theme_linedraw() + ggtitle("Distribution of Maths scores")
```


## Working with ggtheme package


```{r}
ggplot(data=exam_data,
       aes(x=ENGLISH)) + geom_histogram(bins = 20, boundary = 100, color = "grey25", fill= "grey90")+ ggtitle("Distribution of English Scores") + theme_solarized()
```


## Working with hrbthemes package

```{r}
ggplot(data=exam_data,
       aes(x=ENGLISH)) + geom_histogram(bins = 20, boundary = 100, color = "grey25", fill= "grey90")+ ggtitle("Distribution of English Scores") + theme_ipsum_inter()
```


```{r}
ggplot(data=exam_data,
       aes(x=ENGLISH)) + geom_histogram(bins = 20, boundary = 100, color = "grey25", fill= "grey90")+ ggtitle("Distribution of English Scores") + theme_ipsum_inter(axis_title_size = 18, base_size = 15, grid = "Y")
```


# Beyond Single Graph 

```{r}
p1 <- ggplot(data=exam_data, aes(x=MATHS)) + geom_histogram(bins = 20, boundary = 100, color = "grey25", fill = "grey90") + coord_cartesian(xlim=c(0,100)) + ggtitle("Distribution of Maths Score")
p1
```


```{r}
p2 <- ggplot(data=exam_data, aes(x = ENGLISH)) +
  geom_histogram(bins=20, boundary = 100, color="grey25", fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
p2
```


```{r}
p3 <- ggplot(data=exam_data, aes(x= MATHS, y=ENGLISH)) + geom_point() + geom_smooth(method = lm, size=0.5) +  coord_cartesian(xlim = c(0,100), ylim = c(0,100)) + ggtitle("English Scores Vs Maths Scores for Primary 3")
p3
```


## Creating Composite Graphics: pathwork methods

- Tow-Column layout can use **Plus Sign +**
- **Parenthesis ()** are used to create subplot group
- Two-Row layout can use **Devision Sign /**


### Combining two ggplot2 graphs

```{r}
p1 + p2
```


### Combining three ggplot2 graphs

**"|"** is to place the plots beside each other
```{r}
(p1 / p2) | p3
```


### Creating a composite figure with tag

```{r}
((p1 / p2) | p3) + plot_annotation(tag_levels =  'I')
```


### Creating figure with insert

```{r}
p3 + inset_element(p2, left = 0.02, bottom = 0.7, right = 0.5, top = 1)
```


### Creating a composite figure by using patchwork and ggtheme

```{r}
patchwork <- (p1 / p2) | p3 
patchwork & theme_economist()
```
