---
title: "Hands-on Exercise 5"
author: "Min Htet Aung"
description: "This is the Hands-on Exercise 5" 
date: "May 15, 2025"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installing and Launching R Packages

- `igraph`: Powerful package for creating, analyzing, and visualizing complex networks and graph structures.
- `visNetwork`: Enables interactive, customizable network visualizations using `vis.js` in R.
- `lubridate`: Simplifies parsing, handling, and arithmetic of date-time data in R.
- `concaveman`: Generates accurate concave hull polygons around spatial point sets for tight boundary mapping.

```{r}
pacman::p_load(igraph, tidygraph, ggraph, visNetwork, lubridate, clock, tidyverse, graphlayouts, concaveman, ggforce)
```


# Importing Data

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```


## Reviewing the Imported data

```{r}
glimpse(GAStech_edges)
```


## Modifying the data

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate, label = TRUE, abbr = FALSE))
```


## Reviewing the revised data fields

```{r}
glimpse(GAStech_edges)
```


## Wrangling Attributes

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```


## Reviewign the Revised Edges File

```{r}
glimpse(GAStech_edges)
```


# Creating Network Objects using tidygraph

- `tbl_graph()` creates a tbl_graph network object from nodes and edges data.

- `as_tbl_graph()` converts network data and objects to a tbl_graph network. Below are network data and objects supported by `as_tbl_graph()`


## The dplyr verbs in tidygraph

- activate() verb from tidygraph serves as a switch between tibbles for nodes and edges. All dplyr verbs applied to `tbl_graph` object are applied to the active tibble.


## Using `tbl_graph()` to build tidygraph data model

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes, edges = GAStech_edges_aggregated,
                           directed = TRUE)
```


## Reviewing the output tidygraph’s graph object

```{r}
GAStech_graph
```


## Reviewing the output tidygraph’s graph object

- As can be seen from the output above, `GAStech_graph` is a tbl_graph object with 4541 edges and 54 nodes.
- Additionally, the command prints the first three rows of "Edge Data" and the first six rows of "Node Data."
- The Node Data is active, it says. One tibble at a time, the data can be altered thanks to the concept of an active tibble inside a tbl_graph object.


## Changing the Active Object

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```


# Plotting Static Network Graphs with ggraph package

ggraph is an extension of ggplot2, making it easier to carry over basic ggplot skills to the design of network graphs.

As in all network graph, there are three main aspects to a ggraph’s network graph, they are:

- nodes,
- edges and
- layouts.
For a comprehensive discussion of each of this aspect of graph, please refer to their respective vignettes provided.


## Plotting a Basic Network Graph

The code chunk below uses `ggraph()`, `geom-edge_link()` and `geom_node_point()` to plot a network graph by using `GAStech_graph`. Before your get started, it is advisable to read their respective reference guide at least once.

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```


## Changing the default network graph theme

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```


## Changing the Coloring of the plot

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```


## Fruchterman and Reingold layout

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```


## Modifying network nodes

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```


## Modifying edges

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```


# Creating Facet Graphs

- `facet_nodes()` whereby edges are only draw in a panel if both terminal nodes are present here,
- `facet_edges()` whereby nodes are always drawn in al panels even if the node data contains an attribute named the same as the one used for the edge facetting, and
- `facet_graph()` faceting on two variables simultaneously.


## Working with `facet_edges()`

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```


## Working with `facet_edges()`

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```


## A Framed Facet Graph

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```


## Working with `facet_nodes()`

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```


# Network Metrics Analysis

## Computing centrality indices

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```


## Visualising Network Metrics

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```


## Visualising Community

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(
    group_edge_betweenness(
      weights = Weight, 
      directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(
    aes(
      width=Weight), 
    alpha=0.2) +
  scale_edge_width(
    range = c(0.1, 5)) +
  geom_node_point(
    aes(colour = community))  

g + theme_graph()
```


```{r}
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
         betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y, 
        group = community, 
        fill = community),  
    alpha = 0.2,  
    expand = unit(0.3, "cm"),  # Expand
    radius = unit(0.3, "cm")  # Smoothness
  ) + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                      size = betweenness_measure),
                      color = "black",
                      shape = 21)
  
g + theme_graph()
```


# Building Interactive Network Graph with visNetwork

## Data preparation

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```


## Plotting the first interactive network graph

```{r}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```


## Working with layout

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```


## Working with visual attributes - Nodes

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```


```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```


## Working with visual attributes - Edges

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```


## Interactivity

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```



