---
title: "Take-home Exercise 03"
author: "Min Htet Aung"
description: "This is the Take-home Exercise 3" 
date: "Jun 7, 2025"
date-modified: "last-modified"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1. The Task

In this take-home exercise, you are required to select one of the module of your proposed Shiny application and complete the following tasks:

-   To evaluate and determine the necessary R packages needed for your Shiny application are supported in R CRAN,
-   To prepare and test the specific R codes can be run and returned the correct output as expected,
-   To determine the parameters and outputs that will be exposed on the Shiny applications, and
-   To select the appropriate Shiny UI components for exposing the parameters determine above.

# 2. Overview

This project focuses on developing a Shiny application that analyzes a large-scale musical knowledge graph to uncover trends, relationships, and rising stars in the Oceanus Folk genre. By integrating multiple R packages for data manipulation, visualization, and interactive UI, the application enables users to explore artist contributions, genre evolution, and influence networks. Through techniques like heatmaps, PCA clustering, radar charts, and ego networks, the analysis identifies emerging talents, compares artist trajectories, and predicts the future of Oceanus Folk. The project aims to provide an intuitive, data-driven tool that supports music industry insights and talent discovery.

# 3. Loading Packages and Importing Data

## 3.1. Loading Packages

| Packages | Purpose |
|----------------------|--------------------------------------------------|
| `tidyverse` | Core data manipulation and visualization tools (`dplyr`, `ggplot2`, `tibble`, etc.). |
| `jsonlite` | Reads the JSON file into a structured R format for further processing. |
| `SmartEDA` | Automates exploratory data analysis (EDA), summaries, and profiling. |
| `tidygraph` | Enables graph/network operations using tidy data principles. |
| `ggraph` | Visualization extension for `tidygraph`, helps draw static network diagrams with `ggplot2`. |
| `igraph` | Low-level graph manipulation and analysis (e.g., neighbors, components, layout). |
| `lubridate` | Easy handling and manipulation of date/time data. |
| `janitor` | Cleans messy column names and tabular data efficiently. |
| `ggthemes` | Offers extra themes and styles for `ggplot2` visualizations. |
| `plotly` | Creates interactive plots from `ggplot2` (like tooltips in heatmaps and PCA plots). |
| `DT` | Creates interactive HTML data tables for viewing in RMarkdown or Shiny. |
| `ggfortify` | Helps plot PCA and clustering results directly from `prcomp` and `kmeans`. |
| `visNetwork` | Builds interactive network graphs (like EGO networks) in HTML widgets. |
| `factoextra` | Enhances visualization and interpretation of multivariate results (PCA, clustering). |
| `FactoMineR` | Performs multivariate analysis like PCA, MCA, CA—underlying support for `factoextra`. |
| `fmsb` | Creates radar/spider charts—used to compare rising star profiles visually. |
| `forcats` | A helper package for working with categorical variables (factors), especially useful for reordering levels in plots. |

```{r}
pacman::p_load(tidyverse, jsonlite, SmartEDA, tidygraph, ggraph, igraph, lubridate, janitor, ggthemes, plotly, DT, ggfortify, visNetwork, factoextra, FactoMineR, fmsb, forcats)
```

## 3.2. Importing Data

This step reads the knowledge graph data from the JSON file (`MC1_graph.json`) into R as a structured list object for further processing.

```{r}
kg <- fromJSON("data/MC1_graph.json")
```

## 3.3. Inspect structure

This step inspects the high-level structure of the imported JSON object, revealing it contains metadata (`directed`, `multigraph`), graph-level details, and two main components: a ***nodes table with 17,412 entries*** and a ***links (edges) table with 37,857 connections**.*

```{r}
str(kg,max.level = 1)
```

# 4. Data Preparation and Cleaning

This section converts the JSON data into tidy data frames for nodes and edges, cleans column names, inspects the data structure, and ensures that date fields (like release dates) are properly formatted for analysis.

## 4.1. Convert JSON to DataFrames

This step extracts the `nodes` and `links` components from the ***JSON object*** and converts them into separate tibble data frames (`nodes_tbl` and `edges_tbl`) for easier manipulation and analysis.

```{r}
nodes_tbl <- as_tibble(kg$nodes)
edges_tbl <- as_tibble(kg$links)
```

## 4.2. Explore and Clean the Data

This step provides an overview of the structure and contents of the nodes and edges data frames, showing that the dataset includes ***17,412 nodes*** (e.g., songs, people, albums) with attributes like genre and release date, and ***37,857 edges*** representing various relationships such as collaborations or influences.

```{r}
glimpse(nodes_tbl)
glimpse(edges_tbl)
```

## 4.3. Clean and Inspect

This step standardizes the column names in both the nodes and edges tables using `clean_names()` to ensure consistent, analysis-friendly formatting for downstream processing.

```{r}
nodes_tbl <- clean_names(nodes_tbl)
edges_tbl <- clean_names(edges_tbl)
```

## 4.4. Convert Dates

This step converts the `release_date` column in the nodes table from character to integer format, allowing for proper chronological analysis and visualization.

```{r}
nodes_tbl <- nodes_tbl %>%
  mutate(release_date = as.integer(release_date))
```

## 4.5. Creating Graph Object

This step prepares the data for network analysis by constructing a directed graph object. It begins by mapping node IDs to row indices, then remaps the source and target columns in the edges to match these indices. After filtering out incomplete connections, the cleaned nodes and edges are used to build a `tbl_graph` object, enabling efficient graph operations and visualizations using the `tidygraph` framework.

```{r}
# Step 1: Create a mapping from node `id` to row index
id_map <- tibble(id = nodes_tbl$id,
                 index = seq_len(nrow(nodes_tbl)))

# Step 2: Map `source` and `target` in edges to the correct row index
edges_tbl_mapped <- edges_tbl %>%
  left_join(id_map, by = c("source" = "id")) %>%
  rename(from = index) %>%
  left_join(id_map, by = c("target" = "id")) %>%
  rename(to = index)

# Step 3: Remove any edges with missing mapped values
edges_tbl_clean <- edges_tbl_mapped %>%
  filter(!is.na(from), !is.na(to)) %>%
  mutate(from = as.integer(from),
         to = as.integer(to))

# Step 4: Ensure the node table is clean
nodes_tbl_clean <- nodes_tbl %>%
  mutate(id = as.integer(id)) %>%
  distinct()

# Step 5: Create the graph object
graph <- tbl_graph(nodes = nodes_tbl_clean, edges = edges_tbl_clean, directed = TRUE)
```

# 5. Exploratory Data Analysis

the frequency of different edge types in the music knowledge graph, representing the types of relationships between entities (e.g., people, songs, albums). The most common connection is `PerformerOf` with ***13,587*** instances, followed by `ComposerOf`, `LyricistOf`, and `RecordedBy`, highlighting the dominant role of performance and creation-related links in the dataset. Less frequent edge types like `MemberOf` and `DirectlySamples` suggest fewer band affiliations or sampling relationships, reflecting how musical influence is more often conveyed through performance and composition.

## 5.1. Count of Edge Types

This bar chart visualizes the frequency of different edge types in the music knowledge graph, representing the types of relationships between entities (e.g., people, songs, albums). The most common connection is `PerformerOf` with ***13,587*** instances, followed by `ComposerOf`, `LyricistOf`, and `RecordedBy`, highlighting the dominant role of performance and creation-related links in the dataset. Less frequent edge types like `MemberOf` and `DirectlySamples` suggest fewer band affiliations or sampling relationships, reflecting how musical influence is more often conveyed through performance and composition.

```{r}
ggplot(data = edges_tbl, aes(y = edge_type)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), 
            hjust = -0.2, color = "black", size = 2) +  # Data labels
  labs(title = "Distribution of Edge Types",       # Main title
       subtitle = "Count of different connection types",  # Subtitle
       y = "Edge Type",                           # Y-axis label
       x = "Count") +                             # X-axis label
  theme_minimal() +                               # Clean theme
  theme(plot.title = element_text(size = 14, face = "bold"),  # Title formatting
        axis.text.y = element_text(size = 10))
```

## 5.2. Count of Node Types

This bar chart illustrates the distribution of different node types in the music knowledge graph. The most prevalent node type is `Person` ***(11,136 entries)***, indicating a strong focus on individual artists, followed by `Song` and `RecordLabel`. Fewer entries for `Album` and `MusicalGroup` suggest that the dataset emphasizes people and their direct musical contributions over collective entities or album-level data. This breakdown helps frame the scope of the dataset and its emphasis on artist-level relationships.

```{r}
ggplot(data = nodes_tbl, aes(y = node_type)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), 
            hjust = -0.2, color = "black", size = 2.5) +  # Data labels
  labs(title = "Distribution of Node Types",       # Main title
       subtitle = "Count of different node categories",  # Subtitle
       y = "Edge Type",                           # Y-axis label
       x = "Count") +                             # X-axis label
  theme_minimal() +                               # Clean theme
  theme(plot.title = element_text(size = 14, face = "bold"),  # Title formatting
        axis.text.y = element_text(size = 10))
```

## 5.3. Popularity over Time (by Genre)

This heatmap visualizes the number of song releases over time by genre, capturing the evolution and relative activity across musical styles. Each cell represents the number of songs released in a given year for a specific genre, with darker shades indicating higher counts. The visualization helps identify trends, bursts of activity, and the emergence of genres over decades.

```{r}
heatmap_data <- nodes_tbl %>%
  filter(node_type == "Song", !is.na(release_date)) %>%
  count(genre, release_date) %>%
  ungroup() %>%
  na.omit()

# Create ggplot heatmap
p <- ggplot(heatmap_data, 
            aes(x = release_date, 
                y = fct_rev(fct_infreq(genre)), 
                fill = n,
                text = paste0("Genre: ", genre, 
                              "<br>Release Date: ", release_date, 
                              "<br>Number of Songs: ", n))) +
  geom_tile(color = "white", size = 1) +  # Make tiles more distinct
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(name = "# of Songs", 
                      low = "sky blue", 
                      high = "dark blue") +  # Match calendar-style red palette
  labs(x = "Release Year", y = NULL, title = "Song Releases Over Time by Genre") +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 6),
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.position = "top"
  ) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))

# Convert to interactive plot
ggplotly(p, tooltip = "text")
```

::: callout-tip
### Insights

-   Oceanus Folk shows consistent and increasing activity, especially after 2020, aligning with Sailor Shift’s rise.
-   Indie Pop and Indie Folk have maintained strong, steady growth, becoming dominant post-2010.
-   Genres like Dream Pop and Synthwave also show rising popularity after 2015, reflecting global music trends.
-   Post-2020 shows a broader diversification across niche genres, suggesting increased genre blending and experimentation.
:::

## 5.4. Notable Songs and Their Genres

This horizontal bar chart displays the number of notable songs categorized by genre, providing insight into which musical styles have had the most significant impact or recognition. The count is derived from songs marked as notable in the dataset, highlighting genres that have produced widely recognized or influential works.

```{r}
nodes_tbl %>%
  filter(notable == TRUE & node_type == "Song") %>%
  count(genre, sort = TRUE) %>%
  ggplot(aes(x = reorder(genre, n), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = n), hjust = -0.1, size = 2.5) +  # Data labels
  coord_flip() +
  labs(
    title = "Count of Notable Songs by Genre",
    x = "Genre",
    y = "Number of Notable Songs"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.title = element_text(size = 12)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))  # Add space for labels
```

::: callout-tip
### Insights

-   Dream Pop leads with 563 notable songs, indicating strong critical or cultural influence.
-   Indie Folk and Synthwave also have high numbers, showing their sustained popularity.
-   Oceanus Folk has 179 notable songs, signaling its substantial and growing relevance.
:::

# 6. Descriptive Stastistics

## 6.1. Rising Star Profile

This step builds a structured profile for each artist in the graph to identify and compare rising stars in the music industry. It starts by extracting metadata on songs, albums, and people, then maps contribution edges (e.g., PerformerOf, ComposerOf) to link artists to their works. The code calculates key metrics per artist — such as total number of works, notable songs, Oceanus Folk contributions, debut year, and time to achieve notability. It also counts collaborations based on influence-related edges (e.g., CoverOf, InterpolatesFrom).

Finally, all these attributes are combined into a unified `rising_star_profile` table, which is displayed interactively. This profile helps quantify artistic impact and growth trajectory — laying the foundation for identifying standout talent and predicting future Oceanus Folk stars.

::: callout-note
### Puropse

-   To quantify artist trajectories using measurable criteria.
-   To identify fast-rising, influential artists based on productivity, genre focus, and collaboration.
-   To support comparative and predictive analysis in later steps of the challenge.
:::

```{r}
# Step 1: Extract metadata for songs/albums and people
songs_albums <- nodes_tbl_clean %>%
  filter(node_type %in% c("Song", "Album")) %>%
  select(work_id = id, release_date, genre, notable)

people <- nodes_tbl_clean %>%
  filter(node_type == "Person") %>%
  distinct(id, .keep_all = TRUE) %>%
  select(person_id = id, name)

# Step 2: Define edge types
contribution_types <- c("ComposerOf", "PerformerOf", "ProducerOf", "LyricistOf", "RecordedBy")
collab_types <- c("LyricalReferenceTo", "CoverOf", "InterpolatesFrom", "DirectlySamples")

# Step 3: Map contribution edges (index → actual IDs)
created_links <- edges_tbl_clean %>%
  filter(edge_type %in% contribution_types) %>%
  left_join(id_map, by = c("from" = "index")) %>%
  rename(person_id = id) %>%
  left_join(id_map, by = c("to" = "index")) %>%
  rename(work_id = id)

# Step 4: Link artist contributions to songs/albums
artist_works <- created_links %>%
  left_join(songs_albums, by = "work_id") %>%
  filter(!is.na(release_date))  # Drop invalid entries

# Step 5: Compute profile metrics for each artist
rising_star_profile <- artist_works %>%
  group_by(person_id) %>%
  summarise(
    total_works = n(),
    notable_works = sum(notable, na.rm = TRUE),
    oceanus_folk_works = sum(genre == "Oceanus Folk", na.rm = TRUE),
    first_release = min(as.integer(release_date), na.rm = TRUE),
    first_notable = min(as.integer(release_date[notable == TRUE]), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    first_notable = ifelse(is.infinite(first_notable), NA_integer_, first_notable),
    time_to_notability = ifelse(!is.na(first_notable),
                                first_notable - first_release, NA)
  )

# Step 6: Count number of collaboration-type links per artist
collaborations <- edges_tbl_clean %>%
  filter(edge_type %in% collab_types) %>%
  count(from, name = "collaborations") %>%
  left_join(id_map, by = c("from" = "index")) %>%
  rename(person_id = id)

# Step 7: Merge collaboration counts into main profile
rising_star_profile <- rising_star_profile %>%
  left_join(collaborations, by = "person_id") %>%
  mutate(collaborations = replace_na(collaborations, 0))

# Step 8: Add person names and reorder columns
rising_star_profile <- rising_star_profile %>%
  inner_join(people, by = "person_id") %>%
  relocate(person_id, name)

# Step 9: Render interactive data table
datatable(
  rising_star_profile,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE,
    order = list(list(3, 'desc'))  # Sort by total_works by default
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-size: 16px; font-weight: bold;',
    'Table: ', 'Rising Star Profile of Artists'
  )
)

```

# 7. Network Analysis and Visualization

This section constructs and explores the music knowledge graph by mapping relationships between artists, songs, albums, and genres. It transforms the cleaned data into a graph structure using node and edge indices, enabling detailed network analysis. These visualizations help reveal collaboration patterns, influence pathways, and the structural role of key artists—particularly in tracing the spread and evolution of Oceanus Folk.

## 7.1. Influence Network for Oceanus Folk Genre

This step constructs an interactive ***influence network*** focused on contributors to notable Oceanus Folk albums. It identifies artists who composed, performed, or produced such albums and maps their direct connections to albums and other collaborators using a 1-hop ego network.

The network shows ***“Person” nodes in blue*** and ***“Album” nodes in yellow***, enabling exploration of how artists are interconnected through their work in the Oceanus Folk genre.

```{r}
# Step 1: Identify notable Oceanus Folk albums
oceanus_folk_album_ids <- nodes_tbl_clean %>%
  filter(genre == "Oceanus Folk", node_type == "Album", notable == TRUE) %>%
  pull(id)

# Step 2: Define edge type categories
contribution_types <- c("ComposerOf", "PerformerOf", "ProducerOf", "LyricistOf", "RecordedBy")
influence_types <- c("LyricalReferenceTo", "CoverOf", "InterpolatesFrom", "DirectlySamples", "InStyleOf")
relevant_types <- c(contribution_types, influence_types)

# Step 3: Build ID mapping
id_map <- tibble(index = 1:nrow(nodes_tbl_clean), id = nodes_tbl_clean$id)

# Step 4: Get people who contributed to the Oceanus Folk albums
oceanus_contributions <- edges_tbl_clean %>%
  filter(edge_type %in% contribution_types) %>%
  left_join(id_map, by = c("to" = "index")) %>%
  rename(to_id = id) %>%
  filter(to_id %in% oceanus_folk_album_ids)

# Step 5: Identify key artists connected to those albums
oceanus_artist_indices <- unique(oceanus_contributions$from)

# Step 6: Get all 1-hop edges for those artists (contributions + influence)
oceanus_ego_edges <- edges_tbl_clean %>%
  filter((from %in% oceanus_artist_indices | to %in% oceanus_artist_indices) &
           edge_type %in% relevant_types)

# Step 7: Gather node indices involved in those edges
oceanus_ego_node_indices <- unique(c(oceanus_ego_edges$from, oceanus_ego_edges$to))

# Step 8: Filter nodes to only relevant types
included_types <- c("Person", "Album")
oceanus_ego_nodes <- nodes_tbl_clean[oceanus_ego_node_indices, ] %>%
  filter(node_type %in% included_types)

# Step 9: Remap index for visualization
index_lookup <- tibble(
  old_index = oceanus_ego_node_indices,
  new_index = 1:length(oceanus_ego_node_indices)
)

# Step 10: Reindex edges for visNetwork
oceanus_ego_edges_vis <- oceanus_ego_edges %>%
  left_join(index_lookup, by = c("from" = "old_index")) %>%
  rename(from_new = new_index) %>%
  left_join(index_lookup, by = c("to" = "old_index")) %>%
  rename(to_new = new_index) %>%
  filter(!is.na(from_new), !is.na(to_new)) %>%
  transmute(
    from = from_new,
    to = to_new,
    label = edge_type,
    title = paste("Edge Type:", edge_type),
    arrows = "to"
  )

# Step 11: Prepare nodes for visNetwork
oceanus_ego_nodes_vis <- oceanus_ego_nodes %>%
  mutate(
    id = 1:n(),
    label = ifelse(!is.na(name), name, node_type),
    group = case_when(
      id %in% oceanus_folk_album_ids ~ "Oceanus_Folk_Album",
      node_type == "Person" ~ "Person",
      node_type == "Album" ~ "Album",
      TRUE ~ "Other"
    ),
    title = paste0("<b>", name, "</b><br>Type: ", node_type)
  ) %>%
  select(id, label, group, title)

# Step 12: Render the interactive network
visNetwork(oceanus_ego_nodes_vis, oceanus_ego_edges_vis, height = "700px", width = "100%") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visLayout(randomSeed = 42) %>%
  visLegend(useGroups = TRUE, position = "right", main = "Node Types")
```

# 8. Clustering Analysis

## 8.1. Data Preparation

```{r}
# Step 1: Filter only artists with meaningful data
cluster_data <- rising_star_profile %>%
  filter(!is.na(first_release)) %>%
  select(name, total_works, notable_works, oceanus_folk_works, 
         first_release, time_to_notability, collaborations) %>%
  drop_na()  # remove rows with NAs

# Step 2: Scale the data (important for clustering)
cluster_scaled <- cluster_data %>%
  select(-name) %>%
  scale()
```

## 8.2. Determine Optimal Number of Clusters (Elbow Method)

```{r}
set.seed(123)
fviz_nbclust(cluster_scaled, kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal K")
```

## 8.3. Apply K-Means Clustering

```{r}
set.seed(123)
k_result <- kmeans(cluster_scaled, centers = 4, nstart = 25)

# Add cluster labels to your data
cluster_data <- cluster_data %>%
  mutate(cluster = factor(k_result$cluster))
```

## 8.4. Interactive PCA Plot

```{r}
pca_result <- prcomp(cluster_scaled, scale. = TRUE)
pca_data <- as.data.frame(pca_result$x[, 1:2])  # Only PC1 and PC2

# Add cluster info and names
pca_data <- pca_data %>%
  mutate(
    cluster = factor(k_result$cluster),
    name = cluster_data$name  # assuming name column from earlier
  )

pca <- ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(aes(text = paste0(
    "<b>Name:</b> ", name, 
    "<br><b>Cluster:</b> ", cluster
  )), alpha = 0.8, size = 2) +
  stat_ellipse(type = "norm", linetype = 2, aes(group = cluster), show.legend = FALSE) +
  labs(
    title = "Interactive PCA Cluster Plot",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )


ggplotly(pca, tooltip = "text")

```

## 8.5. Cluster Summary Table

```{r}
cluster_summary <- cluster_data %>%
  group_by(cluster) %>%
  summarise(
    total_works = mean(total_works, na.rm = TRUE),
    notable_works = mean(notable_works, na.rm = TRUE),
    oceanus_folk_works = mean(oceanus_folk_works, na.rm = TRUE),
    first_release = mean(first_release, na.rm = TRUE),
    time_to_notability = mean(time_to_notability, na.rm = TRUE),
    collaborations = mean(collaborations, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  )

datatable(cluster_summary, options = list(pageLength = 5))
```

# 9. Visualize the careers of three artists. Compare and contrast their rise in popularity and influence.

This section explores and compares the careers of three selected artists—***Sailor Shift, Beatrice Albright, and Daniel O'Connell***—using multiple visualizations. It analyzes their contributions to songs and albums, highlights key metrics like total works, notable songs, Oceanus Folk involvement, and collaborations, and visualizes their timelines and ego networks. Together, these insights reveal differences in artistic output, genre focus, and influence patterns, helping to understand each artist’s trajectory and role in shaping the Oceanus Folk genre.

## 9.1. Identify the three Artists

This step defines the three focal artists—***Sailor Shift, Beatrice Albright, and Daniel O'Connell***—whose careers will be analyzed and compared in terms of influence, productivity, and genre involvement.

```{r}
# Step 1: Manually select Sailor Shift
artist_sailor <- rising_star_profile %>%
  filter(str_detect(name, regex("Sailor Shift", ignore_case = TRUE)))

# Step 2: Select one of Sailor's ex-bandmates (Ivy Echoes members)
ivy_members <- c("Maya Jensen", "Lila Hartman", "Jade Thompson", "Sophie Ramirez")
artist_peer <- rising_star_profile %>%
  filter(name %in% ivy_members) %>%
  arrange(desc(notable_works)) %>%  # pick the most prominent one
  slice(1)

# Step 3: Select an emerging Oceanus Folk artist (not in Ivy Echoes)
artist_emerging <- rising_star_profile %>%
  filter(!name %in% c(ivy_members, "Sailor Shift")) %>%
  filter(oceanus_folk_works > 0, time_to_notability >= 0) %>%
  arrange(time_to_notability, desc(notable_works)) %>%
  slice(1)

# Combine the three
selected_artists <- bind_rows(artist_sailor, artist_peer, artist_emerging)

# View selected artists
datatable(selected_artists, options = list(pageLength = 5), caption = "Selected Artists for Career Comparison")
```

## 9.2. Getting their Contributions (Songs/Albums)

This step compiles all songs and albums contributed to by the selected artists by linking their IDs to relevant edges in the graph (e.g., `ComposerOf`, `PerformerOf`, `ProducerOf`, `LyricistOf`). It joins these connections with metadata such as release date, genre, and notability, producing a detailed table of each artist's creative works. This forms the foundation for comparing their productivity, genre focus, and the recognition they’ve received across their careers.

```{r}
# Step 0: Prepare artist ID and name list
artist_nodes <- selected_artists %>%
  select(artist_id = person_id, artist_name = name)

# Step 1: Map artist IDs to tidygraph node index
artist_nodes_indexed <- artist_nodes %>%
  left_join(id_map, by = c("artist_id" = "id")) %>%
  rename(index = index)

relevant_edges <- edges_tbl_clean %>%
  filter(edge_type %in% relevant_types)

# Step 4: Match contributions/influences with selected artists (by `from` index)
artist_edges <- relevant_edges %>%
  left_join(artist_nodes_indexed, by = c("from" = "index")) %>%
  filter(!is.na(artist_name))  # Only keep edges linked to selected artists

# Step 5: Join edges with metadata from nodes_tbl_clean
artist_works <- artist_edges %>%
  mutate(to = as.integer(to)) %>%
  left_join(
    nodes_tbl_clean %>%
      select(work_id = id, work_name = name, work_type = node_type,
             release_date, genre, notable),
    by = c("to" = "work_id")
  ) %>%
  filter(work_type %in% c("Song", "Album")) %>%
  select(artist_name, work_name, release_date, genre, notable, work_type, edge_type)

# Step 6: Clean release date format (if needed)
artist_works <- artist_works %>%
  mutate(release_date = as.integer(release_date)) %>%
  arrange(artist_name, release_date)

# Step 7: Display interactive table
datatable(
  artist_works,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE
  ),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-size: 16px; font-weight: bold;',
    'Table: ', 'Works Contributed to or Influenced by Selected Artists'
  )
)

```

## 9.3. Plotting Career Timeline

This timeline chart visualizes the release dates of works by the three selected artists, focusing on the top genres they contributed to. Each point represents a song or album, with shapes indicating genre and colors showing whether the work is notable.

```{r}
top_genres <- artist_works %>%
  count(genre, sort = TRUE) %>%
  slice_head(n = 6) %>%
  pull(genre)

# Build a vector of those genres sorted Z-A
levels_genre_rev <- sort(top_genres, decreasing = TRUE)   # <- key line

# ── 2. Filter artist_works and set genre factor levels in that order 
artist_subset <- artist_works %>%
  filter(genre %in% top_genres) %>%
  mutate(
    genre = factor(genre, levels = levels_genre_rev)  # keeps legend Z-A
  )

# ── 3. Plot 
ggplot(
  artist_subset,
  aes(x = release_date, y = artist_name, shape = genre, color = notable)
) +
  geom_jitter(width = 0.3, height = 0.3, size = 3, alpha = 0.7) +
  scale_color_manual(values = c(`TRUE` = "#a1d99b", `FALSE` = "#fcbba1")) +
  labs(
    title  = "Timelines (Top Genres Sorted Z–A)",
    shape  = "Genre",
    color  = "Notable"
  ) +
  theme_minimal()
```

::: callout-tip
### Insights

-   Sailor Shift has the longest and most consistent timeline, with steady output across genres, especially Oceanus Folk.
-   Daniel O’Connell and Beatrice Albright are active mainly between 2028–2036, with fewer works and less genre diversity.
-   Most of Beatrice’s and Daniel’s works are notable, suggesting quality over quantity.
-   Sailor’s mix of notable and non-notable works shows broad experimentation and prolific output.
:::

## 9.4. Plotting EGO Network Diagram for Top 3 Artists

This interactive ego network visualizes the immediate (1-hop) relationships of the three selected artists—***Sailor Shift, Beatrice Albright, and Daniel O'Connell*****.** It maps their direct connections to other people, songs, albums, groups, and record labels using `visNetwork`, with nodes colored by type and edges labeled by relationship.

```{r fig.height=10, fig.width=12}
# 1. Filter nodes to selected artists (Person type only)
ego_artist_nodes <- nodes_tbl_clean %>%
  filter(node_type == "Person", name %in% selected_artists$name) %>%  
  left_join(id_map, by = c("id" = "id")) %>%
  select(artist_name = name, artist_id = id, index)

ego_indices <- ego_artist_nodes$index  # vector of node indices

# 2. Extract 1-hop ego network from graph object
ego_subgraph <- graph %>%
  activate(nodes) %>%
  filter(node_is_adjacent(ego_indices) | row_number() %in% ego_indices)

# 3. Prepare node data for visNetwork
nodes_data <- ego_subgraph %>%
  activate(nodes) %>%
  as_tibble() %>%
  mutate(
    id = row_number(),
    label = ifelse(!is.na(name), name, node_type),
    group = node_type,
    title = paste0("<b>", label, "</b><br>Type: ", node_type)
  ) %>%
  select(id, label, group, title)

# 4. Prepare edge data for visNetwork
edges_data <- ego_subgraph %>%
  activate(edges) %>%
  as_tibble() %>%
  mutate(
    from = as.integer(from),
    to = as.integer(to),
    label = edge_type,
    title = paste("Edge Type:", edge_type)
  ) %>%
  select(from, to, label, title)

# 5. Plot with visNetwork
visNetwork(nodes_data, edges_data, height = "600px", width = "100%") %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
    nodesIdSelection = TRUE
  ) %>%
  visInteraction(navigationButtons = TRUE) %>%
  visLayout(randomSeed = 42) %>%
  visLegend(
    useGroups = TRUE,
    position = "right",
    main = "Node Type"
  )
```

# 10. Predicting the next Oceanus Folk stars over the next five year

This section identifies promising new artists who are likely to become the next Oceanus Folk stars. Using the rising star profiles, it filters for individuals with a strong commitment to the genre, a notable body of work, and a relatively recent or rapid rise in visibility. By applying data-driven thresholds and ranking by impact (e.g., notable works, collaborations), the analysis highlights top candidates who exhibit the traits of future genre leaders. These predictions help spotlight emerging talent shaping the next wave of Oceanus Folk music.

## 10.1. Defining Thresholds to Filter Candidates

This step applies a set of criteria to identify emerging Oceanus Folk artists with high potential. The filters select artists with at least 5 Oceanus Folk works, 6 or more notable works, a debut within the past 20 years, and either a rapid rise to notability or recent entry. The resulting candidates—***Zacharie Martins, Orla Seabloom, Tao Jin, and Lei Jin***—demonstrate strong genre commitment and early success, positioning them as likely future stars.

```{r}
# Define the current year
current_year <- 2025

# Identify promising Oceanus Folk artists (potential future stars)
future_stars <- rising_star_profile %>%
  filter(
    oceanus_folk_works >= 5,                               # strong Oceanus Folk involvement
    notable_works >= 6,                                    # recognized contributions
    !name %in% selected_artists$name,                      # exclude current analyzed stars
    first_release >= (current_year - 10),                  # debuted in last 10 years
    time_to_notability <= 3 | is.na(time_to_notability)    # either fast-rising or new
  ) %>%
  arrange(desc(notable_works), desc(collaborations)) 

# Preview top candidates
future_stars %>% select(name, total_works, notable_works, oceanus_folk_works, collaborations, time_to_notability)
```

## 10.2. Pick Top 3 Candidates and Justify

This step selects the top 3 rising Oceanus Folk artists based on a combined score of notable works, collaborations, and genre contributions. The selected artists—***Zacharie Martins, Orla Seabloom, and Tao Jin***—stand out for their strong early recognition and dedication to Oceanus Folk. Orla’s higher collaboration count suggests strong network potential, while Zacharie and Tao show consistent output and fast-rising impact, making them strong candidates for future stardom.

```{r}
# Choose the top 3 candidates
top_3_future_stars <- future_stars %>% 
  slice_max(order_by = notable_works + collaborations + oceanus_folk_works, n = 3)

top_3_future_stars
```

## 10.3. Radar Chart for Each Candidates

This radar chart visually compares the top three future Oceanus Folk stars—***Zacharie Martins, Orla Seabloom, and Tao Jin***—across five key dimensions: total works, notable works, Oceanus Folk contributions, collaborations, and rapid rise (time to notability). Each chart normalizes scores to a 0–100 scale, helping highlight each artist’s relative strengths.

```{r}

radar_data <- top_3_future_stars %>%
  select(total_works, notable_works, oceanus_folk_works, collaborations, time_to_notability) %>%
  as.data.frame()

# Set row names to artist names
rownames(radar_data) <- top_3_future_stars$name

# Rename columns for better display
colnames(radar_data) <- c("Total", "Notable", "Oceanus\nFolk", "Collabs", "Rapid\nRise")

# Invert Time to Notability (BEFORE scaling)
radar_data[, "Rapid\nRise"] <- max(radar_data[, "Rapid\nRise"], na.rm = TRUE) - 
                               radar_data[, "Rapid\nRise"]

# Normalize all features to 0-100 range
radar_data_scaled <- as.data.frame(lapply(radar_data, function(x) {
  scales::rescale(x, to = c(0, 100))
}))
rownames(radar_data_scaled) <- rownames(radar_data)

# Set layout with better margins
par(mfrow = c(1, 3), 
    mar = c(2, 2, 4, 2),
    oma = c(1, 1, 3, 1),
    family = "sans")

# Loop and plot
for (i in 1:nrow(radar_data_scaled)) {
  artist_name <- rownames(radar_data_scaled)[i]
  artist_values <- radar_data_scaled[i, , drop = FALSE]
  
  radar_individual <- rbind(
    rep(100, ncol(radar_data_scaled)),
    rep(0, ncol(radar_data_scaled)),
    artist_values
  )
  
  radarchart(radar_individual,
             axistype = 1,
             pcol = "darkorange",
             pfcol = rgb(1, 0.5, 0, 0.3),
             plwd = 2,
             cglcol = "gray",
             cglty = 1,
             cglwd = 0.8,
             axislabcol = "darkblue",
             caxislabels = paste0(seq(0, 100, 25), "%"),
             vlcex = 0.85,
             title = artist_name,
             calcex = 0.8,
             cex.main = 1.2)
}
```
