---
title: "Hands-on Exercise 9.5"
author: "Min Htet Aung"
description: "This is the Hands-on Exercise 9.5" 
date: "Jun 18, 2025"
date-modified: "last-modified"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Treemap Visualisation with R

## Installing and Launching R Packages

```{r}
pacman::p_load(treemap, treemapify, tidyverse) 
```

## Data Wrangling

This dataset provides information of private property transaction records in 2018. The dataset is extracted from REALIS portal (https://spring.ura.gov.sg/lad/ore/login/index.cfm) of Urban Redevelopment Authority (URA).

### Importing the data set

In the code chunk below, read_csv() of readr is used to import realis2018.csv into R and parsed it into tibble R data.frame format.

```{r}
realis2018 <- read_csv("data/realis2018.csv")
```

### Data Wrangling and Manipulation

The realis2018 data frame contains highly disaggregated transaction records, making it unsuitable for direct treemap plotting. To prepare the data for effective visualization, we will perform the following data transformation steps:

Group the data by key categorical variables: Project Name, Planning Region, Planning Area, Property Type, and Type of Sale.

- Summarize transaction details by computing:
  - Total units sold using the sum of No. of Units,
  - Total area using the sum of Area (sqm),
  - Median unit price using the median of Unit Price ($ psm), and
  - Median transacted price using the median of Transacted Price ($).

Grouping affects the verbs as follows:

-   `grouped select()` is the same as `ungrouped select()`, except that grouping variables are always retained.
-   `grouped arrange()` is the same as ungrouped; unless you set `.by_group = TRUE`, in which case it orders first by the grouping variables.
-   `mutate()` and `filter()` are most useful in conjunction with window functions (like `rank()`, or `min(x) == x`). They are described in detail in vignette(“window-functions”).
-   `sample_n()` and `sample_frac()` sample the specified number/fraction of rows in each group.
-   `summarise()` computes the summary for each group.

### Grouped summaries without the Pipe

The code chank below shows a typical two lines code approach to perform the steps.

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`, 
                               `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped, 
                          `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE), 
                          `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

### Grouped summaries with the pipe

The code chunk below shows a more efficient way to tackle the same processes by using the pipe, %\>%:

```{r}
realis2018_summarised <- realis2018 %>% 
  group_by(`Project Name`,`Planning Region`, 
           `Planning Area`, `Property Type`, 
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), 
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

## Designing Treemap with treemap Package

`treemap` package is a R package specially designed to offer great flexibility in drawing treemaps. The core function, namely: `treemap()` offers at least 43 arguments. In this section, we will only explore the major arguments for designing elegent and yet truthful treemaps.

### Designing a static treemap

In this section, `treemap()` of Treemap package is used to plot a treemap showing the distribution of median unit prices and total unit sold of resale condominium by geographic hierarchy in 2017.

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

### Using the basic arguments

The code chunk below designed a treemap by using three core arguments of `treemap()`, namely: index, vSize and vColor.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Working with vColor and type arguments

In the code chunk below, type argument is define as value.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Colours in treemap package

In treemap visualizations, the color scheme is controlled by two key arguments: mapping and palette. The main difference between "value" and "manual" mapping lies in how colors are assigned to the data values.

- In the "value" setting, the color palette is treated as diverging (e.g., ColorBrewer’s "RdYlBu"), where:
  - 0 maps to the middle color (e.g., white or yellow),
  - Negative and positive extremes map to the left-end and right-end colors respectively, using the range from -max(abs(values)) to max(abs(values)).

- In the "manual" setting, colors are mapped more directly:
  - min(values) maps to the left-end color,
  - max(values) maps to the right-end color,
  - The mean of the range maps to the middle color.

### The “value” type treemap

The code chunk below shows a value type treemap.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### The “manual” type treemap

The “manual” type does not interpret the values as the “value” type does. Instead, the value range is mapped linearly to the colour palette.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

Things to learn from the code chunk above:

-   The colour scheme used is very copnfusing. This is because mapping = (min(values), mean(range(values)), max(values)). It is not wise to use diverging colour palette such as RdYlBu if the values are all positive or negative

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Treemap Layout

`treemap()` supports two popular treemap layouts, namely: “squarified” and “pivotSize”. The default is “pivotSize”.

The squarified treemap algorithm (Bruls et al., 2000) produces good aspect ratios, but ignores the sorting order of the rectangles (sortID). The ordered treemap, pivot-by-size, algorithm (Bederson et al., 2002) takes the sorting order (sortID) into account while aspect ratios are still acceptable.

### Working with algorithm argument

The code chunk below plots a squarified treemap by changing the algorithm argument.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### Using sortID

When “pivotSize” algorithm is used, sortID argument can be used to dertemine the order in which the rectangles are placed from top left to bottom right.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## Designing Treemap using treemapify Package

treemapify is a R package specially developed to draw treemaps in ggplot2. In this section, you will learn how to designing treemps closely resemble treemaps designing in previous section by using treemapify.

### Designing a basic treemap

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") + 
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

### Defining hierarchy

Group by Planning Region

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") + 
  geom_treemap()
```

Group by Planning Area

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap()
```

Adding boundary line

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

## Designing Interactive Treemap using d3treeR

### Installing d3treeR package

This slide shows you how to install a R package which is not available in cran.

```{r}
library(devtools)
install_github("timelyportfolio/d3treeR")
library(d3treeR)
```

### Designing An Interactive Treemap

The codes below perform two processes.

treemap() is used to build a treemap by using selected variables in condominium data.frame. The treemap created is save as object called tm.

```{r}
tm <- treemap(realis2018_summarised,
        index=c("Planning Region", "Planning Area"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        title="Private Residential Property Sold, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

Then d3tree() is used to build an interactive treemap.

```{r}
d3tree(tm,rootname = "Singapore" )
```
